Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-07-04 Manuscript.tex
Donna comments

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
& ND-1016-3 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & - Tierra del & ND-1016-4 & analysis \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & & analysis; \\
& & & & & Mg, Li, Sr, Ca. \\
& & & & & \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & Stable oxygen \\
& & & Westray (UK) & QG1-7189-2 & and carbon \\
& & & & QG2-1061-1 & isotope analysis \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

Using this system we carried out elemental imaging of the shell
specimens at a resolution of 50--100 µm distance between sampling spots.
Each spot was irradiated 10 times with the 3 first spectra discarded as
a cleaning step and the remaining spectra summed to get an average Mg/Ca
intensity ratio for each sample spot. Subsequently, we re-sampled the
section using a line scan at 10 µm resolution. This leads to an overlap
between sample spots, but also allows for a continuous record without
gaps. Intensity ratios were filtered for cases with high relative
standard deviation (i.e.~more than 10\%). These occurred in places where
the previous sampling procedures for carbonate powder as part of the
oxygen isotope analysis left an uneven sample surface introducing
variability in the plasma generation and thus uncertainty in the data of
these locations.

@@

High-quality \(\delta\)\textsuperscript{18}O values were acquired from
previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the stable isotope analysis methodology can be found. In short, all
shells were sectioned along the main growth axis to expose the internal
growth structure. From the exposed section carbonate powder samples were
acquired using a micromill with resolutions of 100--200 µm and by
targeting the calcitic M+2 layer of the shells. Analytical precision of
the isotope values was consistently at 0.05--0.10‰ and values are
reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
Belemnite) standard.

@@

\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
shells from Cambaceres Bay.}

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species (\emph{Nacella} sp., and \emph{Patella
vulgata}, respectively. Due to the high sampling resolution, all shells
show seasonally resolved and multi-year quasi-sinusoidal records of SST
change, with \emph{Nacella} sp. shells living 2--3 years and \emph{P.
vulgata} living regularly over 3 years. In those instances where longer
lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b) the temporal-resolution is reduced from more
than a dozen of samples per year, to only around 7.

@@

Because it is not possible to directly associate the exact sample
locations used in the analysis of stable oxygen isotopes with the LIBS
data, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy-sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy data sets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment. The process involved iterative adjustments to minimise the
overall distance between corresponding points in the two data sets.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating and low ratios indicating low temperature
periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
2D-distribution of elemental ratios across the entire section of ORK-LT5
and the preserved anterior side of QG1-7188-1 (additional maps can be
found in the supplementaries). The repeating patterns are found across
the calcitic layers that are exterior to the myostracum (layers m+2 and
m+3) as well as the layers that are interior (m-2)
\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
Mg/Ca intensity ratio consistently falls below the range of interest
(0.2 and above). Compared to other patelloid shells, some of whose
interior is almost entirely made of aragonite, this layer is very thin.
\newline

@@

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
intensity ratios expected based on the stable isotope data and
elemental-imaging. That said, some of the variability of Mg/Ca intensity
ratios within one season such as the summer period of QG1-7188-1 between
0.5 and 2.0 mm distance to the shell edge (Figure~\ref{fig-Pat_LIBS} d),
is visible in a more dramatic manner than it appears in the 2D-image,
showing the downside of line-scanning as opposed to the broader milling
approach used for the analysis of carbonate powder
\citep{Ferguson2011-zl}.

Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
to its \(\delta\)\textsuperscript{18}O-values (similar graphs for other
specimen can be found in the supplementaries). The measurements of the
distance to the shell edge are not entirely identical, most likely
because one growth increment can have a range of distances to the edge,
depending on where one measures, and a line scan along the interior
would be shorter than a line scan along the exterior. That said,
increases and decreases of both proxies seem to mirror each other and to
be well aligned.

@@

Aligning the records programmatically using dynamic time warping,
allowed us to compare them directly to determine specimen-specific
equations and quantify the correlation of both proxies
(Figure~\ref{fig-Pat_Corr}). The equations for all specimens are
different but the majority (barring one: QG1-7246-1) do seem to cluster
around shared parameters (\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065
* Mg/Ca). The various coefficients of determination
(R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
(QG1-7189-2). These specimens are also the oldest and youngest
specimens, respectively, suggesting that lower growth rates and time
averaging had a negative effect on the correlation of both proxies. The
mean R\textsuperscript{2} value of the individual \emph{P. vulgata}
specimens is 0.89, suggesting a generally good fit between the two
proxies.

@@

\subsection{\texorpdfstring{\emph{Nacella}
sp.}{Nacella sp.}}\label{nacella-sp.}

The LIBS data for \emph{Nacella} sp. specimens is less straightforward
with shells being much thinner and thus patterns being more difficult to
make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in
Figure~\ref{fig-Nac_LIBS} (see additional graphs in supplementaries)
also experience increases of Mg/Ca intensity ratio towards the exterior
of the shell, similar to the \emph{P. vulgata} specimens from Orkney.
This intra-increment heterogeneity combined with the thinness of the
shell, further complicates their analysis compared to \emph{P. vulgata}
or other \emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless,
repeating patterns were visible in the anterior section of the shell,
which were also mirrored in the inner layers (m-2) of specimens
NM-1016-3 (Figure~\ref{fig-Nac_LIBS} b).

@@

\caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
analysed \emph{Nacella} sp. specimens. (\emph{a+c}) ND-1016-3 and
(\emph{b}+\emph{d}) NM-1016-3. Note that the ranges of Mg/Ca intensity
ratios are not identical but are chosen arbitrarily to increase
contrast. Areas in grey consist of low-Mg aragonite and fall outside the
chosen colour-range. Note the inset in (\emph{c}) where the shell edge
has been resampled using a higher resolution (50 µm instead of 100 µm)
to better understand the Mg/Ca ratios at the shell edge.}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
emission lines. While the intensity ratio depends chiefly on the chosen
emission lines to calculate the proportion of the chosen Magnesium to
the chosen Calcium peak, the low ratios still indicate lower
concentrations than seen elsewhere, or seen only in aragonitic parts of
e.g.~\emph{Patella} shells. Similar observations have been made before
\citep{Graniero2017-io}. Strong intra-increment heterogeneities are
visible along the exterior of the shells (m+3 layer), where Mg/Ca
intensity ratios are higher, particularly in specimen ND-1016-3
(Figure~\ref{fig-Nac_LIBS}). The step from lower to higher along the
increment is not gradual, as is the case in NM-1016-3 (see inset of
Figure~\ref{fig-Nac_LIBS} b), but rather step-wise. Sampling across this
boundary in a linear scan would most likely lead to erroneous results.
Also sampling across gradual changes as in NM-1016-3 can influence the
overall correlation graph for a specimen.

@@

Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O-values
(similar graphs for the other \emph{Nacella} sp. specimen can be found
in the supplementaries). Compared to the \emph{P. vulgata} records,
these mirror each other much more clearly, due to the brief period
recorded in the shells and its simple make-up.

@@

\caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} sp. specimens.}

@@

The \emph{Nacella} sp. specimens showed similarly clustered
correlations, which interestingly were not grouped by species. In fact,
specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3
(\emph{Nacella magellanica}) are almost identical in their fitted
equations (Figure~\ref{fig-Nac_Corr}). Both also share a high
coefficient of determination (R\textsuperscript{2}) of 0.95 and 0.92
respectively, with the remaining specimens being somewhat lower
(R\textsuperscript{2} = 0.76 for ND-1016-4 and R\textsuperscript{2}=0.84
for NM-1016-3). The mean R\textsuperscript{2} value for each
\emph{Nacella} species is 0.86 for \emph{Nacella deaureata} and 0.88 for
\emph{Nacella magellanica}. Why two specimens of different species are
so well aligned is not entirely certain and randomness cannot for now be
ruled out. That said, since they both have a high R\textsuperscript{2}
value, it might just be that they were both minimally affected by
factors other than SST, including the sampling location of our line
scans.

@@

\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} sp.
specimens}

@@

The Mg/Ca ratios from the 2D elemental imaging present positive results
on SST being the main factor behind Mg/Ca concentration in three
Atlantic limpet species. Annually repeating patterns of increasing
(spring/summer) and decreasing (autum/winter) SST were also visible in
the line scans, albeit less straightforward than in the 2D-data.
Nevertheless, the resulting R\textsuperscript{2}-values for each species
were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
R\textsuperscript{2}}=0.88). These results are similar to previously
published data on \emph{Patella} sp. shells (Table
\hyperref[Table_2]{2}): R\textsuperscript{2} values of \emph{P.
depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca
ratios in increments with rich organic contents \citep{Hausmann2019-fi}.

@@

\emph{P. rustica} has two studied specimens (JL1 and JL2) with
R\textsuperscript{2}-values of 0.02 and 0.8, respectively
\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
from two \emph{P. caerulea} shells (JM00 and JM30 with
R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
used determine a general Mg/Ca-SST relationship with an
R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
calculation, as well as the second year of growth in JM00, which
produced anomalous data. While no additional information about those
specimens is available, it seems likely that discarded the Mg/Ca ratios
from JM00, which were interpreted as an ontogenetic trend, were
potentially influenced by the same intra-increment heterogeneity we see
in some of our shell specimens as well. Since these shells were analysed
by homogenising the carbonate powder of the m+2 layer and the more
exterior m+3 layer, these heterogeneities could potentially increased
the Mg/Ca ratio unpredictably, leading to what might look like an
ontogenetic trend in a linear dataset. Whether these anomalies affected
JL1 in the same way, albeit stronger, is not clear but it might well be
possible.

@@

Another shell whose results we aim to explain further is ORK-LT5, which
has previously been shown to have no substantial correlation with
\(\delta\)\textsuperscript{18}O or SST in implication
\citep{Graniero2017-io}. Different to other specimens of this study, we
were able to resample this specimen and to acquire Mg/Ca data on a
slightly higher resolution (20--30 µm spots in a 100 µm raster compared
to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as
using a continuous high-resolution line-scan at (20--30 µm spots spaced
at 10 µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
previously missed summer peaks at around 2--3 mm distance to the shell
edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
mirroring the change towards warmer temperatures at the time of
collection (August). Including these missing parts lead to a better
alignment of both records and a resulting R\textsuperscript{2}-value of
0.88 (Figure~\ref{fig-ORK_sub}).

@@

This study presents high-resolution and 2D information on Mg/Ca
concentrations in modern and archaeological patelloid shells and their
relationship to sea surface temperature. The results for \emph{Patella
vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
consistently promising with mean R\textsuperscript{2}-values of 0.89,
0.86, and 0.88 respectively.

LIBS elemental imaging provided some insight into elemental anomalies
found particulary in the external parts (m+3 layer) of the shell, which
introduce intra-incremental heterogeneity and which cannot be reconciled
with modern SST changes or SST changes recorded in the
\(\delta\)\textsuperscript{18}O of the same shells. Our 2D approach
allowed us to avoid these areas of intra-increment heterogeneity and
focus on areas that are predominantly controlled by changes in SST.

>>>>

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
& ND-1016-3 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & - Tierra del & ND-1016-4 & analysis \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & & analysis; \\
& & & & & Mg, Li, Sr, Ca. \\
& & & & & \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O \\
& & & Westray (UK) & QG1-7189-2 & and \(\delta\)\textsuperscript{13}C \\
& & & & QG2-1061-1 & analysis \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

Using this system we carried out elemental imaging of the shell
specimens at a resolution of 50--100 µm distance between sampling spots.
Each spot was irradiated 10 times with the first 3 spectra discarded as
a cleaning step and the remaining spectra summed to get an average Mg/Ca
intensity ratio for each sample spot. Subsequently, we re-sampled the
section using a line scan at 10 µm resolution. This leads to an overlap
between sample spots, but also allows for a continuous record without
gaps. Intensity ratios were filtered for cases with high relative
standard deviation (i.e., more than 10\%). These occurred in places
where the previous sampling procedures for carbonate powder as part of
the oxygen isotope analysis left an uneven sample surface introducing
variability in the plasma generation and thus uncertainty in the data of
these locations.

@@

High-quality \(\delta\)\textsuperscript{18}O values were acquired from
previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the methodology for the stable isotope analyses can be found. In short,
all shells were sectioned along the maximum growth axis to expose the
internal growth structure. From the exposed section carbonate powder
samples were acquired using a micromill with resolutions of 100--200 µm
and by targeting the calcitic M+2 layer of the shells. Analytical
precision of the isotope ratios was consistently at 0.05--0.10‰ and
values are reported in per mil units (‰) relative to the VPDB (Vienna
Pee Dee Belemnite) standard.

@@

\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} spp. limpet
shells from Cambaceres Bay.}

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species (\emph{Nacella} spp., and \emph{Patella
vulgata}, respectively). Due to the high-resolution sampling, all shells
show seasonally resolved and multi-year quasi-sinusoidal records of SST
change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
vulgata} living regularly over 3 years. In those instances where longer
lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
reduced from more than a dozen of samples per year, to only around 7.

@@

Because it is not possible to directly associate the exact sample
locations used in the analysis of oxygen isotope ratios with the LIBS
data, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy datasets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment (see code in \citep{Hausmann_2024OSF}). The process involved
iterative adjustments to minimise the overall distance between
corresponding points in the two data sets.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating high temperature periods and low ratios
indicating low temperature periods. As examples,
Figure~\ref{fig-Pat_LIBS} shows the 2D-distribution of elemental ratios
across the entire section of ORK-LT5 and the preserved anterior side of
QG1-7188-1 (additional maps can be found in the supplementaries). The
repeating patterns are found across the calcitic layers that are
exterior to the myostracum (layers m+2 and m+3) as well as the layers
that are interior (m-2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
for a detailed explanations of microstructural layers in \emph{P.
vulgaata}). The layer m-1, which is also interior, consists of aragonite
and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
ratio consistently falls below the range of interest (0.2 and above).
Compared to other patelloid shells, some of whose interior is almost
entirely made of aragonite, this layer is very thin. \newline

@@

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
intensity ratios expected based on the stable isotope data and elemental
imaging. That said, some of the variability of Mg/Ca intensity ratios
within one season such as the summer period of QG1-7188-1 between 0.5
and 2.0 mm distance to the shell edge (Figure~\ref{fig-Pat_LIBS} d), is
visible in a more dramatic manner than it appears in the 2D image. This
shows the downside of line-scanning as opposed to the broader milling
approach used for the analysis of carbonate powder
\citep{Ferguson2011-zl}.

Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
to its \(\delta\)\textsuperscript{18}O values (similar graphs for other
specimen can be found in the supplementaries). The measurements of the
distance to the shell edge are not entirely identical, most likely
because one growth increment can have a range of distances to the edge,
depending on where one measures, and a line scan along the interior
would be shorter than a line scan along the exterior. That said,
increases and decreases of both proxies seem to mirror each other and to
be well aligned.

@@

Aligning the records programmatically using dynamic time warping,
allowed us to compare both proxies for each specimen directly to
determine specimen-specific equations and quantify the correlation of
both proxies (Figure~\ref{fig-Pat_Corr}). The equations for all
specimens are different but the majority (barring one: QG1-7246-1) do
seem to cluster around shared parameters
(\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065 * Mg/Ca). The various
coefficients of determination (R\textsuperscript{2}) range between 0.81
(QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest
and youngest specimens, respectively, suggesting that lower growth rates
and time averaging had a negative effect on the correlation of both
proxies. The mean R\textsuperscript{2} value of the individual \emph{P.
vulgata} specimens is 0.89, suggesting a generally good fit between the
two proxies.

@@

\subsection{\texorpdfstring{\emph{Nacella}
spp.}{Nacella spp.}}\label{nacella-spp.}

The LIBS data for \emph{Nacella} spp. specimens is less straightforward
with shells being much thinner than \emph{P. vulgata} and thus patterns
being more difficult to discern. Both specimens (ND-1016-3 and
NM-1016-3) shown as examples in Figure~\ref{fig-Nac_LIBS} (see
additional graphs in supplementaries) also experience increases of Mg/Ca
intensity ratio towards the exterior of the shell, similar to the
\emph{P. vulgata} specimens from Orkney. This intra-increment
heterogeneity combined with the thinness of the shell, further
complicates their analysis compared to \emph{P. vulgata} or other
\emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless, repeating
patterns were visible in the anterior section of the shell, which were
also mirrored in the inner layers (m-2) of specimens NM-1016-3
(Figure~\ref{fig-Nac_LIBS} b).

@@

\caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
analysed \emph{Nacella} spp. specimens. (\emph{a+c}) ND-1016-3 and
(\emph{b}+\emph{d}) NM-1016-3. Note that the ranges of Mg/Ca intensity
ratios are not identical but are chosen arbitrarily to increase
contrast. Areas in grey consist of low-Mg aragonite and fall outside the
chosen colour-range. Note the inset in (\emph{c}) where the shell edge
has been resampled using a higher resolution (50 µm instead of 100 µm)
to better understand the Mg/Ca ratios at the shell edge.}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} spp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{P. vulgata} above, or 0.5--1.5 in
Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
emission lines. While the intensity ratio depends chiefly on the chosen
emission lines to calculate the proportion of the chosen Magnesium to
the chosen Calcium peak, the low ratios still indicate lower
concentrations than seen elsewhere or seen only in aragonitic parts of
e.g.~\emph{Patella} shells. Similar observations have been made before
\citep{Graniero2017-io}. Strong intra-increment heterogeneities are
visible along the exterior of the shells (m+3 layer), where Mg/Ca
intensity ratios are higher, particularly in specimen ND-1016-3
(Figure~\ref{fig-Nac_LIBS}). The step from lower to higher along the
increment is not gradual, as is the case in NM-1016-3 (see inset of
Figure~\ref{fig-Nac_LIBS} b), but rather stepwise. Sampling across this
boundary in a linear scan would most likely lead to erroneous results.
Also sampling across gradual changes as in NM-1016-3 can influence the
overall correlation graph for a specimen.

@@

Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O values
(similar graphs for the other \emph{Nacella} spp. specimens can be found
in the supplementaries). Compared to the \emph{P. vulgata} records,
these mirror each other much more clearly, due to the brief period
recorded in the shells and thus a lower chance of misaligning annual
extremes.

@@

\caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} spp.
specimens.}

@@

The \emph{Nacella} spp. specimens showed similarly clustered
correlations, which interestingly were not grouped by species. In fact,
specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3 (\emph{N.
magellanica}) are almost identical in their fitted equations
(Figure~\ref{fig-Nac_Corr}). Both also share a high coefficient of
determination (R\textsuperscript{2}) of 0.95 and 0.92 respectively, with
the remaining specimens being somewhat lower (R\textsuperscript{2} =
0.76 for ND-1016-4 and R\textsuperscript{2}=0.84 for NM-1016-3). The
mean R\textsuperscript{2} value for each \emph{N.} species is 0.86 for
\emph{N. deaureata} and 0.88 for \emph{N. magellanica}. Why two
specimens of different species are so well aligned is not entirely
certain and randomness cannot for now be ruled out. That said, since
they both have a high R\textsuperscript{2} value, it might just be that
they were both minimally affected by factors other than SST, including
the sampling location of our line scans.

@@

\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} spp.
specimens}

@@

The Mg/Ca ratios from the 2D elemental imaging present positive results
suggesting that SST is the main factor influencing variations in Mg/Ca
ratios in three Atlantic limpet species. Annually repeating patterns of
increasing (spring/summer) and decreasing (autum/winter) SST were also
visible in the line scans, albeit less straightforward than in the
2D-data. Nevertheless, the resulting R\textsuperscript{2}-values for
each species were high (\emph{P. vulgata mean
R\textsuperscript{2}}=0.89, \emph{N. deaureata mean
R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
R\textsuperscript{2}}=0.88). These results are similar to previously
published data on other \emph{Patella} species shells (Table
\hyperref[Table_2]{2}): R\textsuperscript{2} values of \emph{P.
depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca
ratios in increments with rich organic contents \citep{Hausmann2019-fi}.

@@

\emph{P. rustica} has two studied specimens (JL1 and JL2) with
R\textsuperscript{2}-values of 0.02 and 0.80, respectively
\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
from two \emph{P. caerulea} shells (JM00 and JM30 with
R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
used to determine a general Mg/Ca-SST relationship with an
R\textsuperscript{2} value of 0.79. JL1 was disregarded in that
calculation, as well as the second year of growth in JM00, which
produced anomalous data. While no additional information about those
specimens is available, it seems likely that the discarded Mg/Ca ratios
from JM00, which were interpreted as an ontogenetic trend, were
potentially influenced by the same intra-increment heterogeneity we see
in some of our shell specimens as well. Since these shells were analysed
by homogenising the carbonate powder of the m+2 layer and the more
exterior m+3 layer, these heterogeneities could have potentially
increased the Mg/Ca ratio unpredictably, leading to what might look like
an ontogenetic trend in a linear dataset. Whether these anomalies
affected JL1 in the same way, albeit stronger, is not clear but it might
well be possible.

@@

Another shell whose results we aim to explain further is ORK-LT5, which
has previously been shown to have no substantial correlation with
\(\delta\)\textsuperscript{18}O values or SST by implication
\citep{Graniero2017-io}. In contrast to other specimens in this study,
we were able to resample this specimen and acquire Mg/Ca data on a
slightly higher resolution (20--30 µm spots in a 100 µm raster compared
to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as
using a continuous high-resolution line-scan at (20--30 µm spots spaced
at 10 µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
previously missed summer peaks at around 2--3 mm distance to the shell
edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
mirroring the change towards warmer temperatures at the time of
collection (August). Including these missing parts lead to a better
alignment of both records and a resulting R\textsuperscript{2}-value of
0.88 (Figure~\ref{fig-ORK_sub}).

@@

This study uses LIBS as a novel analytical tool to present
high-resolution and 2D information on Mg/Ca concentrations in modern and
archaeological patelloid shells and their relationship to sea surface
temperature. The results for \emph{P. vulgata}, \emph{N. deaureata}, and
\emph{N. magellanica} were consistently promising with mean
R\textsuperscript{2} values of 0.89, 0.86, and 0.88 respectively.

LIBS elemental imaging provided insight into elemental anomalies found
particulary in the external parts (m+3 layer) of the shell, which
introduce intra-incremental heterogeneity and which cannot be reconciled
with modern SST changes or SST changes recorded in the
\(\delta\)\textsuperscript{18}O values of the same shells. Our 2D
approach allowed us to avoid these areas of intra-increment
heterogeneity and focus on areas that are predominantly controlled by
changes in SST.
